---
title: "Reproducible Research Project Week2"
author: "Aditya Pradana"
date: "January 19, 2017"
output: html_document
---

```{r RR Project Week2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "figure/")
```
  
## R Markdown

This is an R markdown document for Coursera Project in Reproducible Research module in Week 2. This markdown will show the codes to answer the questions in the assignment.


### Loading and preprocessing the data

Below is the code to load data into R and process it.
This is assuming that the dataset file **activity.csv** is in the same folder with R Markdown document. Otherwise, download the data from this link (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).  
Preprocess the data. Convert factor to strings. Convert date string to Date class.

``` {r LoadData}
activity <- read.csv("activity.csv", header=TRUE, stringsAsFactors = FALSE)
activity$date <- as.Date(as.character(activity$date), "%Y-%m-%d")
head(activity)
```

### What is mean total number of steps taken per day?

**1. Total number of steps per day**  
The code below sum the steps, remove NA, group them by date, and name the column in the resulting dataset.
```{r totalStep}
totalsteps <- setNames(aggregate(activity$steps, by=list(activity$date), 
                                  FUN=sum, na.rm=TRUE), 
                                    c("Date", "Steps"))
```


**2. Histogram**  
Code below draws histogram of frequency of total steps and save the plot under ~/figure/ folder.
```{r histogram1}
hist(totalsteps$Steps, 
     xlab="Total Steps Daily", ylab="Frequency", main="Histogram of Total Steps", 
     col="light green")
```


**3. Mean & Median**  
Calculate Mean and Median
```{r MeanMedian1}
##Mean
mean(totalsteps$Steps)

##Median
median(totalsteps$Steps)
```


### What is the average daily activity pattern?

**1. Time series plot**  
Create a dataframe of average steps daily versus interval.  
Name the dataframe column and create the plot.
```{r timeInterval1}
avginterval <- setNames(aggregate(activity$steps, by=list(activity$interval),
                         FUN=mean, na.rm=TRUE), c("Interval", "Average.Steps"))
plot(x=avginterval$Interval, y=avginterval$Average.Steps,
     xlab="Interval", ylab="Average Interval", main="Average steps vs. Interval",
     type="l")

```

**2. Which 5-min interval contains maximum no. of steps**
```{r MaxSteps}
avginterval[avginterval$Average.Steps==max(avginterval$Average.Steps),]
```

We can see that the maximum no. of steps is at interval **835min**.



### Imputing missing values

**1.Calculate number of rows with NA**
```{r NArows}
sum(is.na(activity$steps))
```

**2. Devise strategy for filling in NA**
As the instruction mentions that no sophisticated strategy is needed to fill up NA, it will be filled in using average steps of 5 min interval.

```{r FillNA}
##Fill NA
activity2 <- activity
for (i in 1:length(activity$interval))
{
  if(is.na(activity$steps[i]))
  {
    activity2$steps[i] <- 
      avginterval[which(avginterval$Interval == 
                          activity$interval[i]),2]
  }

}
```

**3.Histogram of total number of steps taken per day with missing data filled in.**  
Create total step dataset which contain sum of steps grouped by date.  
Draw histogram plot.  
Calculate Mean and Median of new dataset with filled missing values.  

```{r histogram2}
totalsteps2 <- setNames(aggregate(activity2$steps, by=list(activity2$date), 
                                    FUN=sum, na.rm=TRUE), 
                                      c("Date", "Steps"))

hist(totalsteps2$Steps, 
     xlab="Total Steps Daily", ylab="Frequency", main="Histogram of Total Steps", 
     col="light blue")

#Mean
mean(totalsteps2$Steps)

#Median
median(totalsteps2$Steps)
```

```{r ImputeValue}
##Imputing missing values with mean of 5min interval makes the histogram looks more like normal distribution graph. The mean shifted to 10766 and since there are 9 similar values around the middle, the median also shifts to that value.
```

  
### Are there differences in activity patterns between weekdays and weekends?

Check the weekdays string and create a column which is a factor of Weekday & Weekend.  
Create a new dataset containing average of 5min interval with filled missing values.  
Draw plot using lattice.
```{r timeInterval2}
## create new column Weektype and fill with Weekday or Weekend based on date
#activity2 <- transform(activity2, Days=weekdays(activity2$date))
for(i in 1:length(activity2$interval))
{
  if(weekdays(activity2$date[i]) %in% c("Saturday", "Sunday"))
  {
      activity2$Weektype[i] <- "Weekend"
  } else {
      activity2$Weektype[i] <- "Weekday"
  }
}
## convert the new column to factor type
activity2$Weektype <- factor(activity2$Weektype, order=TRUE)

## calculate the average 5min interval with filled missing values
avginterval2 <- setNames((aggregate(activity2$steps, by=list(activity2$interval, activity2$Weektype),
                                      FUN=mean, na.rm=TRUE)), 
                                        c("Interval", "Wday", "Steps"))

## Draw plot
library(lattice)
xyplot(Steps ~ Interval | Wday, data=avginterval2, 
       type='l', layout=c(1,2), 
       ylab="Number of Steps")
detach("package:lattice", unload=TRUE)
```

There seems to be a peak in the graph for weekdays. A particular time interval (about 800-900) has a peak in number of steps.

